Split-Apply-Combine with Collapsing Groups

### 2022-12-30 paper/article.Rnw
writing

%% - \Address{} of at least one author
%% - May contain multiple affiliations for each author
%%   (in extra lines, separated by \emph{and}\\).
%% - May contain multiple authors for the same affiliation
%%   (in the same first line, separated by comma).
\Address{
  Mark P.J. van der Loo~\orcidlink{0000-0002-9807-4686}\\
  Research and Development\\
  Statistics Netherlands\\
  Henri Faasdreef 312\\
  2492JP Den Haag, the Netherlands\\
  E-mail: \email{mpj.vanderloo@cbs.nl}\\
  URL: \url{https://www.markvanderloo.eu}\\
  \emph{and}\\
  Leiden Institute of Advanced Computer Science (LIACS)\\
  P.O. Box 9512\\
  2300 RA Leiden, The Netherlands\\
}

@@

In order to use the data frame interface, the input dataset needs to have the
most fine-grained grouping variable. In the running example this is
$A\times B$, so we need to combine that into a single variable, and remove
the other ones.
<<>>=
input$AB <- paste(input$A, input$B, sep="-")
input <- input[-(1:3)]
input
@
We now define the collapsing scheme as follows.
<<>>=
csh <- data.frame(
  AB  = c("1-11", "2-12","2-13","3-21","3-22","3-12")
, AB1 = c("1-1" , "2-1" ,"2-1" ,"3-2" ,"3-2" ,"3-1" )
, A   = c("1"   , "2"   ,"2"   ,"3"   ,"3"   ,"3"   ))
@
The calls to \code{cumulate()} and \code{accumulate()} now look as follows.
<<>>=
accumulate(input, collapse=csh, test = min_records(3), sum)
cumulate(input, collapse=csh, test=min_records(3), tY=sum(Y), tY2 = sum(Y2))
@

@@

\bibliography{refs}

>>>>

%% - \Address{} of at least one author
%% - May contain multiple affiliations for each author
%%   (in extra lines, separated by \emph{and}\\).
%% - May contain multiple authors for the same affiliation
%%   (in the same first line, separated by comma).
\Address{
  Mark P.J. van der Loo~\orcidlink{0000-0002-9807-4686}\\
  Research and Development\\
  Statistics Netherlands\\
  Henri Faasdreef 312\\
  2492JP Den Haag, the Netherlands\\
  E-mail: \email{mpj.vanderloo@cbs.nl}\\
  URL: \url{https://www.markvanderloo.eu}\\
  \emph{and}\\
  Leiden Institute of Advanced Computer Science (LIACS)\\
  University of Leiden\\
  P.O. Box 9512\\
  2300 RA Leiden, The Netherlands\\
}

@@

In order to use the data frame interface, the input dataset needs to have the
most fine-grained grouping variable. In the running example this is
$A\times B$, so we need to combine that into a single variable, and remove
the other ones.
<<>>=
input$AB <- paste(input$A, input$B, sep="-")
input <- input[-(1:3)]
input
@
We now define the collapsing scheme as follows.
<<>>=
csh <- data.frame(
  AB  = c("1-11", "2-12","2-13","3-21","3-22","3-12")
, AB1 = c("1-1" , "2-1" ,"2-1" ,"3-2" ,"3-2" ,"3-1" )
, A   = c("1"   , "2"   ,"2"   ,"3"   ,"3"   ,"3"   ))
@
In this data frame, two consecutive columns should be read as a child-parent
relation. For example, in the first collapsing step the groups defined by
\code{AB == "2-12"} and \code{AB=="2-13"} both collapse to \code{AB1 == "2-1"}.
In this artificial example the codes do not mean anything, but in realistic
cases where codes represent a (hierarchical) classification, domain experts
usually have a good grasp of which codes can be combined.

The calls to \code{cumulate()} and \code{accumulate()} now look as follows.
<<>>=
accumulate(input, collapse=csh, test = min_records(3), sum)
cumulate(input, collapse=csh, test=min_records(3), tY=sum(Y), tY2 = sum(Y2))
@

@@

\newpage
\bibliography{refs}
